---
title: "Project 1"
format: html
editor: visual
---

# Heart disease risk factors

## Intro

In this report we will look at a dataset that contains information about a group of patients undergoing an aniography (a type of X-ray to look at the blood vessels around the heart) where the researches collected several different attributes about the patients and also recorded the

## Methods

## Data exploration

```{r}
# Loading libraries
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r}
# Set directory to to where the data is saved
setwd("C:/Julie/epfl/MATH-517/test_project1")

# Read dataset from csv file
heart_data <- read.csv("heart_disease.csv")

# Remove unwanted columns (e.g., 'cholesterol' and 'age')
heart_data <- heart_data %>%
  select(-trestbps, -fbs, -oldpeak, -thal) 

# Remove the rows with NA values
heart_data <- heart_data %>%
  filter(!is.na(ca))

# Create a new column 'heart_disease' based on binary values
heart_data <- heart_data %>%
  mutate(bin = ifelse(num == 0, 0, 1))  # 0 for no heart disease, 1 for heart disease

heart_data$bin <- as.factor(heart_data$bin)  # Convert bin to a factor
heart_data$num <- as.factor(heart_data$num)  # Convert num to a factor

table(heart_data$bin)


```

Here we are taking a look at the stucture and content of the dataframe in order to get an idea of characteristics of the data.

```{r}
# Get the size of the dataframe(df)
print(paste("The dataset has", nrow(heart_data), "rows and", ncol(heart_data), "columns."))
```

```{r}
# Look the df
head(heart_data)
```

```{r}
table(heart_data$sex)
```

Here is a discription of the data

| Variable Name | Type        | Description                                                                                            |
|------------------|------------------|-------------------------------------|
| age           | Integer     | Age in years                                                                                           |
| sex           | Categorical | Sex (0 = female; 1 = male)                                                                             |
| cp            | Categorical | Chest pain type (1 = typical angina; 2 = atypical angina; 3 = non-anginal pain; 4 = asymptomatic)      |
| chol          | Integer     | Serum cholestoral (mg/dl)                                                                              |
| restecg       | Categorical | Resting electrocardiographic results (0 = normal)                                                      |
| thalach       | Integer     | Maximum heart rate achieved                                                                            |
| exang         | Categorical | Exercise induced angina (0 = no; 1 = yes)                                                              |
| slope         | Categorical | The slope of the peak exercise ST segment (1 = upsloping; 2 = flat; 3 = downsloping)                   |
| ca            | Integer     | Number of major vessels (0-3) colored by fluoroscopy                                                   |
| num           | Integer     | Diagnosis of heart disease (0 = no heart disease; 1 = mild; 2 = moderate; 3 = severe; 4 = very severe) |
| bin           | Categorical | Diagnosis of heart disease (0 = no heart disease; 1 = heart disease present)                           |

```{r}
# Assuming 'df' is your dataframe
df_summary <- heart_data %>%
  group_by(bin) %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE)) %>%
  pivot_longer(cols = -bin, names_to = "variable", values_to = "mean") %>%
  pivot_wider(names_from = bin, values_from = mean, names_prefix = "")

# View the result
print(df_summary)
```

```{r}
ggplot(heart_data, aes(x = age, y = thalach, color = bin)) +
  geom_point(size = 3) +  # Makes the points larger
  geom_smooth(method = "lm", se = FALSE) +  # Add regression lines without confidence intervals
  labs(title = "Scatterplot of Age vs Heart Rate by Heart Disease Status",
       x = "Age",
       y = "Heart Rate",
       color = "Heart Disease Status") +  # Legend title
  theme_minimal()

```

The 'num' columns contains the information about the persons heart disease on a scale of integers from 0 (no presence) to 4.

Be careful about conclutions. Look like there are less people

```{r}
ggplot(heart_data, aes(x = bin, y = age)) +
  geom_violin(trim = FALSE) +
  labs(title = "Violin Plot of Age by Category",
       x = "Category",
       y = "Age") +
  theme_minimal()
```

Women have less angina than men.

```{r}
# Modify the ggplot to use interaction between num and sex
ggplot(heart_data, aes(x = interaction(bin, sex), y = cp)) +
  geom_violin(trim = FALSE) +
  labs(title = "Violin Plot of CP by Combined Categories",
       x = "Combined Category (num and sex)",
       y = "CP") +
  theme_minimal()
```

```{r}
# Convert 'sex' to a factor for labeling
heart_data$sex <- factor(heart_data$sex, levels = c(0, 1), labels = c("Female", "Male"))

# Create the plot with percentages by gender
ggplot(heart_data, aes(x = sex, fill = factor(num))) +
  geom_bar(position = "fill") +  # 'fill' will normalize the bars to percentages
  scale_y_continuous(labels = scales::percent) +  # Format y-axis as percentages
  scale_fill_discrete(name = "Heart Disease Severity", 
                      labels = c("No Heart Disease", "Mild", "Moderate", "Severe", "Very Severe")) +
  labs(title = "Percentage of Heart Disease Severity by Gender", 
       x = "Gender", 
       y = "Percentage") +
  theme_minimal() +
  theme(legend.title = element_blank())

```

```{r}
# Create the plot
ggplot(heart_data, aes(x = chol, fill = bin)) +
  geom_histogram(binwidth = 10, color = "black", alpha = 0.7) +  # Histogram with specified bin width
  facet_wrap(~ bin, scales = "fixed") +  # Ensure both plots share the same y-axis scale
  labs(title = "Comparison of Cholesterol Levels by Heart Disease Status",
       x = "Cholesterol Level (mg/dL)",
       y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "none")  # Remove the legend as we are already faceting

```

0: A lot around 200. 1: more above 200.

We expected bigger differnce. Maybe bias in dataset?

## Referenses

The data:

Janosi, A., Steinbrunn, W., Pfisterer, M., & Detrano, R. (1989). Heart Disease \[Dataset\]. UCI Machine Learning Repository. <https://doi.org/10.24432/C52P4X.>

info about the study:

<https://pubmed.ncbi.nlm.nih.gov/2756873/>
